A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-23 paper.tex
moved signal strength into expt design section and swapped participant to subject

Within the context of visual inference, with \(K\) independent
observers, the visual \(p\)-value can be seen as the probability of
having as many or more subjects detect the data plot than the observed
result.

The approach used in \citet{majumder_validation_2013} is as follows.
Define \(X_j = \{0,1\}\) to be a Bernoulli random variable measuring
whether subject \(j\) detected the data plot, and
\(X = \sum_{j=1}^{K}X_j\) be the total number of observers who detected
the data plot. Then, by imposing a relatively strong assumption that all
\(K\) evaluations are fully independent, under \(H_0\)
\(X \sim \mathrm{Binom}_{K,1/m}\). Therefore, the \(p\)-value of a
lineup of size \(m\) evaluated by \(K\) observer is estimated with
\(P(X \geq x) = 1 - F(x) + f(x)\), where \(F(.)\) is the binomial
cumulative distribution function, \(f(.)\) is the binomial probability
mass function and \(x\) is the realization of number of observers
choosing the data plot.

As pointed out by \citet{vanderplas2021statistical}, this basic binomial
model is deficient. It does not take into account the possible
dependencies in the visual test due to repeated evaluations of the same
lineup, or account for when subjects are offered the option to select
one or more ``most different'' plots, or none, from a lineup. They
suggest three common lineup scenarios: (1) \(K\) different lineups are
shown to \(K\) subjects, (2) \(K\) lineups with different null plots but
the same data plot are shown to \(K\) subjects, and (3) the same lineup
is shown to \(K\) subjects. Scenario 3 is the most feasible to apply,
but has the most dependencies to accommodate for the \(p\)-value
calculation. For Scenario 3, VanderPlas et al propose modeling the
probability of plot \(i\) being selected from a lineup as \(\theta_i\),
where \(\theta_i \sim Dirichlet(\alpha)\) for \(i=1,...,m\) and
\(\alpha > 0\). The number of times plot \(i\) being selected in \(K\)
evaluations is denoted as \(c_i\). In case subject \(j\) makes multiple
selections, \(1/s_j\) will be added to \(c_i\) instead of one, where
\(s_j\) is the number of plots subject \(j\) selected for \(j=1,...K\).
This ensures \(\sum_{i}c_i=K\). Since we are only interested in the
selections of the data plot \(i\), the marginal model can be simplified
to a beta-binomial model and thus the visual \(p\)-value is given as

@@

The power of a visual test on the other hand, may additionally depend on
the ability of the particular subject, as the skill of each individual
may affect the number of observers who identify the data plot from the
lineup \citep{majumder_validation_2013}. To address this issue,
\citet{majumder_validation_2013} models the probability of subject \(j\)
correctly picking the data plot from lineup \(l\) using a mixed-effect
logistic regression, with subjects treated as random effect. Then, the
estimated power of a visual test evaluated by a single subject is the
predicted value obtained from the mixed effects model. However, this
mixed effects model does not work with scenario where subjects are asked
to select one or more most different plots. In this scenario, having the
probability of a subject \(j\) correctly picking the data plot from a
lineup \(l\) is insufficient to determine the power of a visual test
because it does not provide information about the number of selections
made by the subject for the calculation of the \(p\)-value (See Equation
\ref{eq:pvalue-beta-binomial-approx}). Therefore, we directly estimate
the probability of a lineup being rejected by assuming that individual
skill has negligible effect on the variation of the power. This
assumption is not necessary true, but it helps to simplify the model
structure, thereby obviating a costly large-scale experiment to estimate
complex covariance matrices. The same model given in Equation
\ref{eq:logistic-regression-1-1} is applied to model the power of a
visual test.

@@

Our experiment was conducted over three data collection periods to
investigate the difference between conventional hypothesis testing and
visual inference in the application of linear regression diagnostics.
Two types of departures, non-linearity and heteroskedasticity, were
collected during data collection periods I and II. The data collection
period III was designed primarily to measure human responses to null
lineups so that the parameter \(\alpha\) in Equation
\ref{eq:pvalue-beta-binomial} can be estimated. Additional lineups for
both non-linearity and heteroskedasticity, using uniform fitted value
distributions, were included for additional data, and to avoid
participant frustration of too many difficult tasks. Overall, we
collected 7974 evaluations on 1152 unique lineups performed by 443
subjects throughout three data collection periods.

@@

Data collection period I was designed to study the ability of human
subjects to detect the effect of a non-linear term \(\boldsymbol{z}\)
constructed using Hermite polynomials on random vector
\(\boldsymbol{x}\) formulated as

@@

Figure \ref{fig:example-poly-lineup} is one of the non-linearity lineups
used in experiment. This data plot is produced by the non-linearity
model with \(j = 6\), and is in location \(2^3+1\). All five subjects
who viewed the lineup chose this plot as most different, thus all
detected the data plot.

@@

\caption{One of the lineups containing heteroskedasticity pattern used in data collection period II. Can you spot the most different plot? The data plot is positioned at $3^3 - 3^2$}\label{fig:example-heter-lineup}
\end{figure}

An example lineup of this model used in data collection period II is
shown in Figure \ref{fig:example-heter-lineup} with \(a = -1\). The data
plot location is in position \(2^4 + 2\). Nine out of 11 subjects chose
this plot from the lineup.

@@

@@

\hypertarget{controlling-the-strength-of-the-signal}{%
\subsubsection{Controlling the strength of the
signal}\label{controlling-the-strength-of-the-signal}}

As summarised in Table \ref{tab:model-factor-table}, three additional
parameters \(n\), \(\sigma\) and \(b\) are used to control the strength
of the signal so that different difficulty levels of lineups are
generated, and therefore, the estimated power curve will be smooth and
continuous. Parameter \(\sigma \in \{0.5, 1, 2, 4\}\) and
\(b \in \{0.25, 1, 4, 16, 64\}\) are used in data collection periods I
and II respectively. Figure \ref{fig:different-sigma} and
\ref{fig:different-b} demonstrate the impact of these two parameters. A
large value of \(\sigma\) will increase the variation of the error of
the non-linearity model and decrease the visibility of the visual
pattern. The parameter \(b\) controls the standard deviation of the
error across the support of the predictor. Given \(x \neq a\), a larger
value of \(b\) will lead to a larger ratio of the variance at \(x\) to
the variance at \(x - a = 0\), making the visual pattern more obvious.

Three different sample sizes are used (n = 50, 100, 300) in all three
data collection periods. It can be observed from Figure
\ref{fig:different-n} that with fewer data points drawn in a residual
plot, the visual pattern is more difficult to be detected.

@@

\hypertarget{subject-allocation}{%
\subsubsection{Subject allocation}\label{subject-allocation}}

As shown in Table \ref{tab:model-factor-table}, there are a total of
\(4 \times 4 \times 3 \times 4 = 192\) and
\(3 \times 5 \times 3 \times 4 = 180\) number of combinations of
parameter values for non-linearity model and heteroskedasticity model
respectively. Three replications are made for each of the combination
results in \(192 \times 3 = 576\) and \(180 \times 3 = 540\) lineups. In
addition, each lineup is designed to be evaluated by five different
subjects. After attempting some pilot studies internally, we decide to
present a block of 20 lineups to every subject. And to ensure the
quality of the survey data, two lineups with obvious visual patterns are
included as attention checks. Thus, \(576 \times 5 / (20-2) = 160\) and
\(540 \times 5 / (20-2) = 150\) subjects are recruited to satisfy the
design of the data collection period I and II respectively.

@@

All lineups consist of only null plots are planned to be evaluated by 20
subjects. However, presenting only these lineups to subjects are
considered to be bad practices as subjects will lose interest quickly.
Therefore, we plan to collect 6 more evaluations on the 279 lineups with
uniform fitted value distribution, result in
\((36 \times 20 + 279 \times 3 \times 6) / (20-2) = 133\) subjects
recruited for data collection period III.

@@

Subjects for all three data collection periods are recruited from an
crowdsourcing platform called Prolific \citep{palan2018prolific}.
Prescreening procedure is applied during the recruitment, subjects are
required to be fluent in English, with \(98\%\) minimum approval rate
and 10 minimum submissions in other studies.

During the experiment, every subject is presented with a block of 20
lineups. A lineup consists of a randomly placed data plot and 19 null
plots, which are all residual plots drawn with raw residuals on the
y-axis and fitted values on the x-axis. An additional horizontal red
line is added at \(y = 0\) as a helping line.

@@

Surprisingly, the visual test rejects 1 of the 33 (3\%) of lineups where
the conventional test does not reject. Figure \ref{fig:heter-example}
shows this lineup. The data plot in position seventeen displays a
relatively strong heteroskedasticity pattern, and has a strong effect
size (\(log_e(E)=4.02\)). This is reflected by the visual test
\(p\text{-value} = 0.026\). But the BP test \(p\text{-value} = 0.056\),
is slightly above the significance cutoff of \(0.05\). This lineup was
evaluated by 11 subjects, it has experimental factors \(a = 0\)
(``butterfly'' shape), \(b = 64\) (large variance ratio), \(n = 50\)
(small sample size), and a uniform distribution for the fitted values.
It may have been the small sample size and the presence of a few
outliers that may have resulted in the lack of detection by the
conventional test.

>>>>

Within the context of visual inference, with \(K\) independent
observers, the visual \(p\)-value can be seen as the probability of
having as many or more participants detect the data plot than the
observed result.

The approach used in \citet{majumder_validation_2013} is as follows.
Define \(X_j = \{0,1\}\) to be a Bernoulli random variable measuring
whether participant \(j\) detected the data plot, and
\(X = \sum_{j=1}^{K}X_j\) be the total number of observers who detected
the data plot. Then, by imposing a relatively strong assumption that all
\(K\) evaluations are fully independent, under \(H_0\)
\(X \sim \mathrm{Binom}_{K,1/m}\). Therefore, the \(p\)-value of a
lineup of size \(m\) evaluated by \(K\) observer is estimated with
\(P(X \geq x) = 1 - F(x) + f(x)\), where \(F(.)\) is the binomial
cumulative distribution function, \(f(.)\) is the binomial probability
mass function and \(x\) is the realization of number of observers
choosing the data plot.

As pointed out by \citet{vanderplas2021statistical}, this basic binomial
model is deficient. It does not take into account the possible
dependencies in the visual test due to repeated evaluations of the same
lineup, or account for when participants are offered the option to
select one or more ``most different'' plots, or none, from a lineup.
They suggest three common lineup scenarios: (1) \(K\) different lineups
are shown to \(K\) participants, (2) \(K\) lineups with different null
plots but the same data plot are shown to \(K\) participants, and (3)
the same lineup is shown to \(K\) participants. Scenario 3 is the most
feasible to apply, but has the most dependencies to accommodate for the
\(p\)-value calculation. For Scenario 3, VanderPlas et al propose
modeling the probability of plot \(i\) being selected from a lineup as
\(\theta_i\), where \(\theta_i \sim Dirichlet(\alpha)\) for
\(i=1,...,m\) and \(\alpha > 0\). The number of times plot \(i\) being
selected in \(K\) evaluations is denoted as \(c_i\). In case participant
\(j\) makes multiple selections, \(1/s_j\) will be added to \(c_i\)
instead of one, where \(s_j\) is the number of plots participant \(j\)
selected for \(j=1,...K\). This ensures \(\sum_{i}c_i=K\). Since we are
only interested in the selections of the data plot \(i\), the marginal
model can be simplified to a beta-binomial model and thus the visual
\(p\)-value is given as

@@

The power of a visual test on the other hand, may additionally depend on
the ability of the particular participant, as the skill of each
individual may affect the number of observers who identify the data plot
from the lineup \citep{majumder_validation_2013}. To address this issue,
\citet{majumder_validation_2013} models the probability of participant
\(j\) correctly picking the data plot from lineup \(l\) using a
mixed-effect logistic regression, with participants treated as random
effect. Then, the estimated power of a visual test evaluated by a single
participant is the predicted value obtained from the mixed effects
model. However, this mixed effects model does not work with scenario
where participants are asked to select one or more most different plots.
In this scenario, having the probability of a participant \(j\)
correctly picking the data plot from a lineup \(l\) is insufficient to
determine the power of a visual test because it does not provide
information about the number of selections made by the participant for
the calculation of the \(p\)-value (See Equation
\ref{eq:pvalue-beta-binomial-approx}). Therefore, we directly estimate
the probability of a lineup being rejected by assuming that individual
skill has negligible effect on the variation of the power. This
assumption is not necessary true, but it helps to simplify the model
structure, thereby obviating a costly large-scale experiment to estimate
complex covariance matrices. The same model given in Equation
\ref{eq:logistic-regression-1-1} is applied to model the power of a
visual test.

@@

Our experiment was conducted over three data collection periods to
investigate the difference between conventional hypothesis testing and
visual inference in the application of linear regression diagnostics.
Two types of departures, non-linearity and heteroskedasticity, were
collected during data collection periods I and II. The data collection
period III was designed primarily to measure human responses to null
lineups so that the parameter \(\alpha\) in Equation
\ref{eq:pvalue-beta-binomial} can be estimated. Additional lineups for
both non-linearity and heteroskedasticity, using uniform fitted value
distributions, were included for additional data, and to avoid
participant frustration of too many difficult tasks. Overall, we
collected 7974 evaluations on 1152 unique lineups performed by 443
participants throughout three data collection periods.

@@

Data collection period I was designed to study the ability of
participants to detect the effect of a non-linear term
\(\boldsymbol{z}\) constructed using Hermite polynomials on random
vector \(\boldsymbol{x}\) formulated as

@@

Figure \ref{fig:example-poly-lineup} is one of the non-linearity lineups
used in experiment. This data plot is produced by the non-linearity
model with \(j = 6\), and is in location \(2^3+1\). All five
participants who viewed the lineup chose this plot as most different,
thus all detected the data plot.

@@

\caption{One of the lineups containing heteroskedasticity pattern used in data collection period II. Can you spot the most different plot? The data plot is positioned at $3^3 - 3^2$.}\label{fig:example-heter-lineup}
\end{figure}

An example lineup of this model used in data collection period II is
shown in Figure \ref{fig:example-heter-lineup} with \(a = -1\). The data
plot location is in position \(2^4 + 2\). Nine out of 11 participants
chose this plot from the lineup.

@@

\hypertarget{controlling-the-signal-strength-of-the-signal}{%
\subsubsection{Controlling the signal strength of the
signal}\label{controlling-the-signal-strength-of-the-signal}}

The three parameters \(n\), \(\sigma\) and \(b\) are important for
controlling the strength of the signal to generate lineups with a
variety difficulty levels. This will ensure that estimated power curves
will be smooth and continuous, and that participants are allocated a set
of lineups with a range of difficulty.

Parameter \(\sigma \in \{0.5, 1, 2, 4\}\) and
\(b \in \{0.25, 1, 4, 16, 64\}\) are used in data collection periods I
and II respectively. A large value of \(\sigma\) will increase the
variation of the error of the non-linearity model and decrease the
visibility of the visual pattern. The parameter \(b\) controls the
variation in the standard deviation of the error across the support of
the predictor. A larger value of \(b\) generates a larger ratio between
it's smallest and highest values, making the visual pattern more
obvious. The sample size \(n\) sharpens (or blurs) a pattern when it is
larger (or smaller). Figures \ref{fig:different-sigma},
\ref{fig:different-b} and \ref{fig:different-n} demonstrate the impact
of these parameters on signal strength.

@@

The lineups are allocated to participants in a manner that covers the
experimental design and is relatively uniform. To manage this, we use
effect size to measure the signal strength, which helps in assigning a
set of lineups with a range of difficulties to each participant.

@@

\hypertarget{lineup-allocation-to-participants}{%
\subsubsection{Lineup allocation to
participants}\label{lineup-allocation-to-participants}}

As shown in Table \ref{tab:model-factor-table}, there are a total of
\(4 \times 4 \times 3 \times 4 = 192\) and
\(3 \times 5 \times 3 \times 4 = 180\) number of combinations of
parameter values for non-linearity model and heteroskedasticity model
respectively. Three replications are made for each of the combination
results in \(192 \times 3 = 576\) and \(180 \times 3 = 540\) lineups. In
addition, each lineup is designed to be evaluated by five different
participants. After attempting some pilot studies internally, we decide
to present a block of 20 lineups to every participant. And to ensure the
quality of the survey data, two lineups with obvious visual patterns are
included as attention checks. Thus, \(576 \times 5 / (20-2) = 160\) and
\(540 \times 5 / (20-2) = 150\) participants are recruited to satisfy
the design of the data collection period I and II respectively.

@@

All lineups consist of only null plots are planned to be evaluated by 20
participants. However, presenting only these lineups to participants are
considered to be bad practices as participants will lose interest
quickly. Therefore, we plan to collect 6 more evaluations on the 279
lineups with uniform fitted value distribution, result in
\((36 \times 20 + 279 \times 3 \times 6) / (20-2) = 133\) participants
recruited for data collection period III.

@@

Participants for all three data collection periods are recruited from an
crowdsourcing platform called Prolific \citep{palan2018prolific}.
Prescreening procedure is applied during the recruitment, participants
are required to be fluent in English, with \(98\%\) minimum approval
rate and 10 minimum submissions in other studies.

During the experiment, every participant is presented with a block of 20
lineups. A lineup consists of a randomly placed data plot and 19 null
plots, which are all residual plots drawn with raw residuals on the
y-axis and fitted values on the x-axis. An additional horizontal red
line is added at \(y = 0\) as a helping line.

@@

Surprisingly, the visual test rejects 1 of the 33 (3\%) of lineups where
the conventional test does not reject. Figure \ref{fig:heter-example}
shows this lineup. The data plot in position seventeen displays a
relatively strong heteroskedasticity pattern, and has a strong effect
size (\(log_e(E)=4.02\)). This is reflected by the visual test
\(p\text{-value} = 0.026\). But the BP test \(p\text{-value} = 0.056\),
is slightly above the significance cutoff of \(0.05\). This lineup was
evaluated by 11 participants, it has experimental factors \(a = 0\)
(``butterfly'' shape), \(b = 64\) (large variance ratio), \(n = 50\)
(small sample size), and a uniform distribution for the fitted values.
It may have been the small sample size and the presence of a few
outliers that may have resulted in the lack of detection by the
conventional test.
